---
title: "Bayesian causal SEM with ordinal indicators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian causal SEM with ordinal indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(dcvar)
```

`dcvar_causal_sem()` fits four causal structural equation models with Stan.
Each row contains one independent person. Each model contains two treatment
groups and one latent variable with three ordinal indicators.

| Model | Latent variable | Effects |
| --- | --- | --- |
| `latent_covariate` | Covariate | ATE and treatment interaction |
| `latent_outcome` | Outcome | ATE and treatment interaction |
| `latent_mediator` | Mediator after treatment | ATE, ADE, and AIE |
| `latent_mediator_baseline` | Mediator before treatment | ATE, ADE, and AIE |

ATE is the average treatment effect. ADE is the average direct effect under
the adjustment used in the SEM study. AIE is ATE minus ADE. A causal
interpretation requires the assumptions of the study design. The Bayes
estimator does not establish those assumptions from the data.

## Fit the four models

The simulator returns data, true parameters, true effects, and a list of fit
arguments. A scalar `n` gives the total sample size. A vector gives the control
and treatment group sizes. The simulator generates data on the identified
factor scale. Its fit arguments set `standardize = FALSE`.

```{r simulate}
model_names <- c("latent_covariate", "latent_outcome",
                 "latent_mediator", "latent_mediator_baseline")
simulations <- setNames(lapply(model_names, function(model) {
  simulate_dcvar_causal_sem(n = c(100, 100), model = model, seed = 42)
}), model_names)
lapply(simulations, function(sim) sim$effects)
```

The following complete workflow fits each model. RStan is the default backend.
Set `backend = "cmdstanr"` to use an installed CmdStan toolchain.

```{r fit, eval=FALSE}
fits <- lapply(simulations, function(sim) {
  do.call(dcvar_causal_sem, c(sim$args, list(
    seed = 42, chains = 4, iter_warmup = 1000, iter_sampling = 1000
  )))
})
lapply(fits, causal_effects)
lapply(fits, dcvar_diagnostics)
```

For your own data, assign columns to model roles. A latent role names the
factor instead of a data column. `indicators` maps that name to three columns.
The columns must be ordered factors with three to five declared levels.
Retain the same levels in both groups, including levels with no observations
in one group. Data preparation reports missing values and role conflicts.
It does not remove persons or categories.

```{r explicit}
sim <- simulations$latent_covariate
prepared <- prepare_causal_sem_data(
  data = sim$data, model = "latent_covariate", treatment = "x",
  control = 0, treated = 1, outcome = "y", covariate = "xi",
  indicators = list(xi = c("u1", "u2", "u3")), standardize = FALSE
)
prepared$meta$n_group
prepared$stan_data$K
```

The latent outcome model uses a factor name for `outcome` and a column for
`covariate`. Both mediation models require `mediator`, `mediator_baseline`, and
`outcome_baseline`. The latent role is `mediator` in the first mediation model
and `mediator_baseline` in the second.

```{r roles}
lapply(simulations, function(sim) sim$roles)
```

## Measurement and scale

The measurement model uses an ordinal probit likelihood. Loadings and
thresholds are equal across groups. The first loading is one. The marginal
factor mean in the control group is zero. Control item residual SDs are one.
Treatment item residual SDs are free. Each endogenous factor has a free
structural residual variance.

The factor unit comes from these restrictions. It is not the unit of a sum
score. It does not require the simulation loadings or thresholds. A comparison
with effects from another parameterization needs a common factor scale.

By default, the fit standardizes manifest continuous columns across both
groups. It stores the center and scale for each role in `fit$meta$scales`.
`causal_effects()` returns effects in the input outcome units. Interactions
also use the input covariate units. Latent outcomes use the identified factor
unit. `coef()` and parameter summaries use the internal model scale.

## Define the target population

The default target weights are `c(0.5, 0.5)`. They assign equal weight to the
control and treatment populations. This target matches the balanced study
design. It need not match the sample proportions in another data set.

```{r effects, eval=FALSE}
fit <- fits$latent_covariate
causal_effects(fit)
causal_effects(fit, target_weights = c(0.7, 0.3))
causal_effects(fit, covariate_values = c(-1, 0, 1))
effect_draws <- causal_effects(fit, summary = FALSE)
posterior::as_draws_matrix(effect_draws)[1:4, ]
```

Changing weights changes the target population. It does not fit the model
again. The effect function combines group parameters within each posterior
draw. It retains the dependence between all parameters in a draw.

For mediation, ADE adjusts both groups at the same weighted baseline means
and the same weighted unconditional mediator mean. ATE integrates each group's
mediator regression at the common baseline means. AIE is their difference.
With group-specific paths, a product of two path means does not give these
effects. ADE is also different from a natural direct effect.

## Priors and checks

The default priors use normal distributions for paths, intercepts, and free
loadings. SDs have positive normal priors. Treatment item SDs have lognormal
priors. Ordered thresholds have normal priors restricted to their order.
The baseline correlation uses an LKJ prior. See `?prepare_causal_sem_data` for
the parameter names and defaults.

Use prior simulation to examine the implied outcomes and category counts.
You can change the prior parameters through `priors`.

```{r prior}
prior_sim <- simulate_dcvar_causal_sem(
  n = c(100, 100), model = "latent_outcome", parameter_prior = TRUE,
  priors = list(beta_sd = 0.5), seed = 12
)
lapply(prior_sim$data[c("u1", "u2", "u3")], table)
```

Prior draws can generate a constant indicator. The simulator retains that
draw. The data contract rejects globally constant indicators. A calibration
study must count rejected data sets and must not replace them silently.

Inspect `dcvar_diagnostics()` before interpreting effects. The output includes
R-hat, effective sample size, divergences, tree depth, energy diagnostics,
and a separate table for the effects. `incomplete_diagnostics` flags unavailable
values. A small headline R-hat does not clear this flag.
Increase sampling when the Monte Carlo
error is too large. Investigate persistent divergences or poor mixing.

The sampler conditions the person factor on continuous outcomes when present.
It then applies the ordinal likelihood. This exact normal factorization reduces
dependence during sampling. Prior sampling uses the structural factor prior.
`fit$sampling$factor_parameterization` records the method.

```{r predictions, eval=FALSE}
fit <- fits$latent_mediator
dcvar_diagnostics(fit)
predict(fit, type = "outcome", summary = TRUE)
item_draws <- predict(fit, type = "indicators")
latent_draws <- predict(fit, type = "latent")
```

Replicates draw new person factors. They preserve each model's measurement
structure. The outcome model conditions on the supplied manifest covariate.
The latent mediator model conditions on the supplied baselines. The latent
mediator baseline model conditions on the outcome baseline and generates a new
mediator baseline and mediator. The latent covariate model generates both the
covariate and the outcome.

`predict()` returns integer codes for indicator replicates. Use
`fit$meta$levels` to recover category labels. Compare category proportions and
joint response patterns with the observed data. The current prediction method
uses the fit data. It does not accept new persons.

The four models assume normal structural distributions. Copulas in the
original study generate data for tests of misspecification. They are not extra
likelihood options in this interface. Development scripts in
`data-raw/causal-sem` run calibration, recovery, and a separate VITA pilot.
A future LOO method must integrate out the person factor for each complete
person. Conditional item likelihoods are not a substitute for that calculation.
