---
title: "dcVar: Complete Feature Walkthrough"
subtitle: "Testing all package functionality end-to-end"
author: "Benedikt Lugauer"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-summary: "Show code"
execute:
  cache: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| cache: false
library(dcVar)
library(ggplot2)

# Keep cmdstanr CSV outputs in a stable directory for the full render.
cmdstanr_output_dir <- file.path(tempdir(), "cmdstanr-vignette")
dir.create(cmdstanr_output_dir, recursive = TRUE, showWarnings = FALSE)
options(cmdstanr_output_dir = cmdstanr_output_dir)

# Shared MCMC settings (minimal for demo speed)
CHAINS    <- 2
WARMUP    <- 500
SAMPLING  <- 500
SEED      <- 42
T_DEMO    <- 80
```

These MCMC settings are intentionally light so the walkthrough remains
renderable. For real analyses, use more chains and iterations and treat any
`Rhat > 1.01` or low ESS in the printed diagnostics as a signal to retune.

## 1. Trajectory Generators

All seven built-in generators produce rho trajectories of length `T - 1` for use
with `simulate_dcVar()`.

```{r}
#| label: trajectories
trajs <- list(
  constant      = rho_constant(T_DEMO, rho = 0.5),
  decreasing    = rho_decreasing(T_DEMO, rho_start = 0.7, rho_end = 0.3),
  increasing    = rho_increasing(T_DEMO, rho_start = 0.3, rho_end = 0.7),
  random_walk   = rho_random_walk(T_DEMO, sigma_omega = 0.05, seed = SEED),
  step          = rho_step(T_DEMO, rho_before = 0.7, rho_after = 0.3, breakpoint = 0.5),
  double_step   = rho_double_step(T_DEMO, rho_levels = c(0.7, 0.3, 0.7))
)

# Verify all return numeric vectors of length T - 1
stopifnot(all(vapply(trajs, length, integer(1)) == T_DEMO - 1))

# Named scenarios
rho_scenario("decreasing", T = T_DEMO)
rho_scenario("double_relapse", T = T_DEMO)
```

```{r}
#| label: plot-trajectories
set.seed(SEED)
plot_trajectories(T = T_DEMO)
```


## 2. Data Simulation

### 2.1 Normal Margins (default)

```{r}
#| label: sim-normal
sim_norm <- simulate_dcVar(
  T = T_DEMO,
  rho_trajectory = rho_decreasing(T_DEMO, rho_start = 0.7, rho_end = 0.3),
  seed = SEED
)
str(sim_norm, max.level = 1)
head(sim_norm$Y_df)
```

### 2.2 Exponential Margins

```{r}
#| label: sim-exp
sim_exp <- simulate_dcVar(
  T = T_DEMO,
  rho_trajectory = rho_decreasing(T_DEMO, rho_start = 0.7, rho_end = 0.3),
  margins = "exponential",
  skew_direction = c(1, -1),
  seed = SEED
)
str(sim_exp$true_params, max.level = 1)
```

### 2.3 Skew-Normal Margins

```{r}
#| label: sim-sn
sim_sn <- simulate_dcVar(
  T = T_DEMO,
  rho_trajectory = rho_decreasing(T_DEMO, rho_start = 0.7, rho_end = 0.3),
  margins = "skew_normal",
  skew_params = list(alpha = c(3, -2)),
  seed = SEED
)
str(sim_sn$true_params, max.level = 1)
```

### 2.4 Gamma Margins

```{r}
#| label: sim-gamma
sim_gam <- simulate_dcVar(
  T = T_DEMO,
  rho_trajectory = rho_decreasing(T_DEMO, rho_start = 0.7, rho_end = 0.3),
  margins = "gamma",
  skew_direction = c(1, -1),
  skew_params = list(shape = 4),
  seed = SEED
)
str(sim_gam$true_params, max.level = 1)
```

### 2.5 Breakpoint Data

```{r}
#| label: sim-breakpoint
sim_single <- simulate_breakpoint_data(T = T_DEMO, type = "single",
                                        rho_before = 0.7, rho_after = 0.3,
                                        seed = SEED)
sim_double <- simulate_breakpoint_data(T = T_DEMO, type = "double",
                                        rho_levels = c(0.7, 0.3, 0.7),
                                        seed = SEED)
```

### 2.6 Multilevel Data

```{r}
#| label: sim-multilevel
sim_ml <- simulate_dcVar_multilevel(
  N = 10, T = 60,
  rho = 0.5,
  seed = SEED
)
str(sim_ml, max.level = 1)
head(sim_ml$data)
```

### 2.7 SEM Data

```{r}
#| label: sim-sem
sim_sem <- simulate_dcVar_sem(
  T = 60, J = 3,
  seed = SEED
)
str(sim_sem, max.level = 1)
head(sim_sem$data)
```


## 3. DC-VAR Model (Normal Margins) — Full Workflow

### 3.1 Fit the Model

```{r}
#| label: fit-dcvar
#| results: hide
fit_dc <- dcVar(
  sim_norm$Y_df,
  vars = c("y1", "y2"),
  chains = CHAINS,
  iter_warmup = WARMUP,
  iter_sampling = SAMPLING,
  seed = SEED,
  refresh = 0
)
```

### 3.2 Inspect

```{r}
#| label: inspect-dcvar
print(fit_dc)
summary(fit_dc)
coef(fit_dc)
```

### 3.3 Extract

```{r}
#| label: extract-dcvar
# Rho trajectory with credible intervals
rho_df <- rho_trajectory(fit_dc)
head(rho_df)

# VAR parameter summaries
var_params(fit_dc)

# Raw posterior draws (first 5 columns, 5 rows)
d <- draws(fit_dc, format = "draws_matrix")
d[1:5, 1:5]
```

### 3.4 Plot

```{r}
#| label: plot-rho
plot_rho(fit_dc, true_rho = sim_norm$true_params$rho)
```

```{r}
#| label: plot-phi
plot_phi(fit_dc)
```

```{r}
#| label: plot-diag
plot(fit_dc, type = "diagnostics")
```

### 3.5 Diagnostics

```{r}
#| label: diagnostics-dcvar
dcVar_diagnostics(fit_dc)
```

```{r}
#| label: plot-ppc
plot_ppc(fit_dc)
```

### 3.6 PIT Diagnostics

```{r}
#| label: pit-dcvar
pit_vals <- pit_values(fit_dc)
head(pit_vals)

pit_test(fit_dc)

plot_pit(fit_dc)
```

### 3.7 Predictions

```{r}
#| label: predict-dcvar
# Fitted values (one-step-ahead)
fit_vals <- fitted(fit_dc)
head(fit_vals)

# Predictions with confidence intervals
pred <- predict(fit_dc, ci_level = 0.95)
head(pred)
```

### 3.8 Tidy Output

```{r}
#| label: tidy-dcvar
tidy_df <- as.data.frame(fit_dc)
head(tidy_df)
```

### 3.9 Interpretation

```{r}
#| label: interpret-dcvar
interpret_rho_trajectory(fit_dc)
```


## 4. DC-VAR with Non-Gaussian Margins

### 4.1 Exponential Margins

```{r}
#| label: fit-exp-model
#| results: hide
fit_exp <- dcVar(
  sim_exp$Y_df,
  vars = c("y1", "y2"),
  margins = "exponential",
  skew_direction = c(1, -1),
  chains = CHAINS,
  iter_warmup = WARMUP,
  iter_sampling = SAMPLING,
  seed = SEED,
  refresh = 0
)
```

```{r}
#| label: fit-exp
plot_rho(fit_exp, true_rho = sim_exp$true_params$rho)
pit_test(fit_exp)
```

### 4.2 Skew-Normal Margins

```{r}
#| label: fit-sn-model
#| results: hide
fit_sn <- dcVar(
  sim_sn$Y_df,
  vars = c("y1", "y2"),
  margins = "skew_normal",
  chains = CHAINS,
  iter_warmup = WARMUP,
  iter_sampling = SAMPLING,
  seed = SEED,
  refresh = 0
)
```

```{r}
#| label: fit-sn
plot_rho(fit_sn, true_rho = sim_sn$true_params$rho)
pit_test(fit_sn)
```

### 4.3 Gamma Margins

```{r}
#| label: fit-gamma-model
#| results: hide
fit_gam <- dcVar(
  sim_gam$Y_df,
  vars = c("y1", "y2"),
  margins = "gamma",
  skew_direction = c(1, -1),
  chains = CHAINS,
  iter_warmup = WARMUP,
  iter_sampling = SAMPLING,
  seed = SEED,
  refresh = 0
)
```

```{r}
#| label: fit-gamma
plot_rho(fit_gam, true_rho = sim_gam$true_params$rho)
pit_test(fit_gam)
```


## 5. HMM Copula Model

```{r}
#| label: sim-hmm-data
sim_step <- simulate_dcVar(
  T = T_DEMO,
  rho_trajectory = rho_step(T_DEMO, rho_before = 0.7, rho_after = 0.3, breakpoint = 0.5),
  seed = SEED
)
```

### 5.1 Fit

```{r}
#| label: fit-hmm
#| results: hide
fit_hmm <- dcVar_hmm(
  sim_step$Y_df,
  vars = c("y1", "y2"),
  K = 2,
  chains = CHAINS,
  iter_warmup = WARMUP,
  iter_sampling = SAMPLING,
  seed = SEED,
  refresh = 0
)
```

### 5.2 Inspect

```{r}
#| label: inspect-hmm
print(fit_hmm)
summary(fit_hmm)
coef(fit_hmm)
```

### 5.3 HMM-Specific Outputs

```{r}
#| label: hmm-states
states <- hmm_states(fit_hmm)
str(states, max.level = 1)
```

```{r}
#| label: plot-hmm-states
plot_hmm_states(fit_hmm)
```

```{r}
#| label: plot-hmm-transition
plot(fit_hmm, type = "transition")
```

### 5.4 Plot and Diagnostics

```{r}
#| label: plot-hmm-rho
plot_rho(fit_hmm, true_rho = sim_step$true_params$rho)
```

```{r}
#| label: hmm-diagnostics
dcVar_diagnostics(fit_hmm)
pit_test(fit_hmm)
```

### 5.5 Interpretation

```{r}
#| label: interpret-hmm
interpret_rho_trajectory(fit_hmm)
```


## 6. Constant Copula Model

```{r}
#| label: sim-const-data
sim_const <- simulate_dcVar(
  T = T_DEMO,
  rho_trajectory = rho_constant(T_DEMO, rho = 0.5),
  seed = SEED
)
```

### 6.1 Fit

```{r}
#| label: fit-constant
#| results: hide
fit_con <- dcVar_constant(
  sim_const$Y_df,
  vars = c("y1", "y2"),
  chains = CHAINS,
  iter_warmup = WARMUP,
  iter_sampling = SAMPLING,
  seed = SEED,
  refresh = 0
)
```

### 6.2 Inspect

```{r}
#| label: inspect-constant
print(fit_con)
summary(fit_con)
coef(fit_con)
```

### 6.3 Plot

```{r}
#| label: plot-constant-phi
plot_phi(fit_con)
```

```{r}
#| label: plot-constant-diag
plot(fit_con, type = "diagnostics")
```

### 6.4 Interpretation

```{r}
#| label: interpret-constant
interpret_rho_trajectory(fit_con)
```


## 7. Model Comparison

Compare the three core models on the same decreasing-rho data from Section 3.

```{r}
#| label: fit-comparison-models
#| results: hide
# Fit HMM and constant on the same decreasing-rho data
fit_cmp_hmm <- dcVar_hmm(
  sim_norm$Y_df,
  vars = c("y1", "y2"),
  K = 2,
  chains = CHAINS,
  iter_warmup = WARMUP,
  iter_sampling = SAMPLING,
  seed = SEED,
  refresh = 0
)

fit_cmp_con <- dcVar_constant(
  sim_norm$Y_df,
  vars = c("y1", "y2"),
  chains = CHAINS,
  iter_warmup = WARMUP,
  iter_sampling = SAMPLING,
  seed = SEED,
  refresh = 0
)
```

### 7.1 LOO-CV

```{r}
#| label: loo-compare
loo_dc  <- loo::loo(fit_dc)
loo_hmm <- loo::loo(fit_cmp_hmm)
loo_con <- loo::loo(fit_cmp_con)

dcVar_compare(dcVar = fit_dc, hmm = fit_cmp_hmm, constant = fit_cmp_con)
```


## 8. Multilevel Model

### 8.1 Simulate Panel Data

```{r}
#| label: sim-multilevel-fit
sim_ml <- simulate_dcVar_multilevel(
  N = 10, T = 60,
  rho = 0.5,
  seed = SEED
)
```

### 8.2 Fit

```{r}
#| label: fit-multilevel
#| results: hide
fit_ml <- dcVar_multilevel(
  sim_ml$data,
  vars = c("y1", "y2"),
  id_var = "id",
  time_var = "time",
  chains = CHAINS,
  iter_warmup = WARMUP,
  iter_sampling = SAMPLING,
  seed = SEED,
  refresh = 0
)
```

### 8.3 Inspect

```{r}
#| label: inspect-multilevel
print(fit_ml)
summary(fit_ml)
coef(fit_ml)
```

### 8.4 Extract

```{r}
#| label: extract-multilevel
var_params(fit_ml)

# Unit-specific random effects
re <- random_effects(fit_ml)
head(re)

# Rho trajectory (constant for multilevel)
rho_trajectory(fit_ml)
```

### 8.5 Plot

```{r}
#| label: plot-multilevel
plot_random_effects(fit_ml)
```

```{r}
#| label: plot-multilevel-diag
plot(fit_ml, type = "diagnostics")
```


## 9. SEM Model

### 9.1 Simulate Latent Data

```{r}
#| label: sim-sem-fit
sim_sem <- simulate_dcVar_sem(
  T = 60, J = 3,
  seed = SEED
)
```

### 9.2 Fit

```{r}
#| label: fit-sem
#| results: hide
fit_sem <- dcVar_sem(
  sim_sem$data,
  indicators = list(
    latent1 = paste0("y1_", 1:3),
    latent2 = paste0("y2_", 1:3)
  ),
  J = 3,
  lambda = sim_sem$true_params$lambda,
  sigma_e = sim_sem$true_params$sigma_e,
  chains = CHAINS,
  iter_warmup = WARMUP,
  iter_sampling = SAMPLING,
  seed = SEED,
  refresh = 0
)
```

### 9.3 Inspect

```{r}
#| label: inspect-sem
print(fit_sem)
summary(fit_sem)
coef(fit_sem)
```

### 9.4 Extract

```{r}
#| label: extract-sem
# Estimated latent states
ls <- latent_states(fit_sem)
head(ls)

var_params(fit_sem)
```

### 9.5 Plot

```{r}
#| label: plot-sem-latent
plot_latent_states(fit_sem, true_states = sim_sem$latent_states)
```

```{r}
#| label: plot-sem-rho
plot(fit_sem, type = "rho")
```

```{r}
#| label: plot-sem-diag
plot(fit_sem, type = "diagnostics")
```


## 10. Data Preparation (Standalone)

These functions prepare data for Stan without fitting. Useful for inspecting
what gets passed to the model.

```{r}
#| label: data-prep
# DC-VAR data
d_dc <- prepare_dcVar_data(sim_norm$Y_df, vars = c("y1", "y2"))
str(d_dc, max.level = 1)

# HMM data
d_hmm <- prepare_hmm_data(sim_norm$Y_df, vars = c("y1", "y2"), K = 2)
str(d_hmm, max.level = 1)

# Constant data
d_con <- prepare_constant_data(sim_norm$Y_df, vars = c("y1", "y2"))
str(d_con, max.level = 1)

# Multilevel data
d_ml <- prepare_multilevel_data(sim_ml$data, vars = c("y1", "y2"),
                                 id_var = "id", time_var = "time")
str(d_ml, max.level = 1)

# SEM data
d_sem <- prepare_sem_data(sim_sem$data,
                           indicators = list(
                             latent1 = paste0("y1_", 1:3),
                             latent2 = paste0("y2_", 1:3)
                           ),
                           J = 3, lambda = sim_sem$true_params$lambda,
                           sigma_e = sim_sem$true_params$sigma_e)
str(d_sem, max.level = 1)
```


## 11. Parameter Recovery Metrics

```{r}
#| label: recovery-metrics
# Rho recovery from the dcVar fit (Section 3)
# Extract rho trajectory draws for metrics
rho_df <- rho_trajectory(fit_dc)
rho_met <- compute_rho_metrics(
  rho_true = sim_norm$true_params$rho,
  rho_est = rho_df$mean,
  rho_lower = rho_df$q2.5,
  rho_upper = rho_df$q97.5
)
rho_met

# Scalar parameter recovery
vp <- var_params(fit_dc)
phi_met <- compute_param_metrics(
  true_value = sim_norm$true_params$Phi[1, 1],
  est_mean = vp$Phi$mean[1],
  est_lower = vp$Phi$q2.5[1],
  est_upper = vp$Phi$q97.5[1]
)
phi_met
```

### 11.1 Mini Simulation Study

```{r}
#| label: mini-study-fits
#| results: hide
metrics_list <- lapply(1:3, function(rep) {
  sim_i <- simulate_dcVar(
    T = T_DEMO,
    rho_trajectory = rho_decreasing(T_DEMO, rho_start = 0.7, rho_end = 0.3),
    seed = rep * 100
  )
  fit_i <- dcVar(
    sim_i$Y_df,
    vars = c("y1", "y2"),
    chains = 2,
    iter_warmup = 250,
    iter_sampling = 250,
    seed = rep,
    refresh = 0
  )
  rho_df_i <- rho_trajectory(fit_i)
  list(rho = compute_rho_metrics(
    rho_true = sim_i$true_params$rho,
    rho_est = rho_df_i$mean,
    rho_lower = rho_df_i$q2.5,
    rho_upper = rho_df_i$q97.5
  ))
})
```

```{r}
#| label: mini-study
aggregate_metrics(metrics_list)
```


## 12. Utility Functions

```{r}
#| label: utilities
# Bundled Stan source files for supported model variants
data.frame(
  model = c("dcVar", "constant", "hmm", "dcVar", "constant", "hmm", "multilevel", "sem"),
  margins = c("normal", "normal", "normal", "exponential", "gamma", "skew_normal", "normal", "normal"),
  stan_file = basename(c(
    dcVar_stan_path("dcVar"),
    dcVar_stan_path("constant"),
    dcVar_stan_path("hmm"),
    dcVar_stan_path("dcVar", margins = "exponential"),
    dcVar_stan_path("constant", margins = "gamma"),
    dcVar_stan_path("hmm", margins = "skew_normal"),
    dcVar_stan_path("multilevel"),
    dcVar_stan_path("sem")
  ))
)
```

## Session Info

```{r}
#| label: session-info
#| cache: false
sessionInfo()
```
