---
title: "Getting Started with dcVar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with dcVar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(Sys.getenv("NOT_CRAN"), "true")
has_cmdstan <- requireNamespace("cmdstanr", quietly = TRUE) &&
  !is.null(tryCatch(cmdstanr::cmdstan_path(), error = function(e) NULL))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = NOT_CRAN && has_cmdstan
)
```

## Installation

The dcVar package uses [CmdStan](https://mc-stan.org/users/interfaces/cmdstan) for
Bayesian model fitting. You need to install **cmdstanr** and **CmdStan** first:

```{r install, eval = FALSE}
# Install cmdstanr from R-universe (not on CRAN)
install.packages("cmdstanr",
  repos = c("https://stan-dev.r-universe.dev", getOption("repos"))
)

# Install the CmdStan backend
cmdstanr::install_cmdstan()

# Install dcVar from GitHub
install.packages("remotes")
remotes::install_github("benlug/dcvar")
```

## Simulating Data

dcVar includes trajectory generators and a data simulator for testing:

```{r simulate}
library(dcVar)

# Generate a decreasing rho trajectory (therapy effect)
rho_true <- rho_decreasing(T = 150, rho_start = 0.7, rho_end = 0.3)

# Simulate bivariate VAR(1) data with this trajectory
sim <- simulate_dcVar(
  T = 150,
  rho_trajectory = rho_true,
  seed = 42
)

head(sim$Y_df)
```

## Fitting the DC-VAR Model

The main fitting function is `dcVar()`. It accepts a data frame and the names
of two variables to model:

```{r fit}
fit <- dcVar(
  data = sim$Y_df,
  vars = c("y1", "y2"),
  chains = 4,
  iter_warmup = 1000,
  iter_sampling = 2000,
  seed = 123
)
```

## Inspecting Results

```{r inspect}
# Quick overview
print(fit)

# Detailed summary
summary(fit)

# Posterior means for VAR parameters
coef(fit)

# Time-varying rho as a data frame
rho_df <- rho_trajectory(fit)
head(rho_df)

# MCMC diagnostics
dcVar_diagnostics(fit)
```

### Fitted Values and Predictions

The `fitted()` method returns one-step-ahead fitted values from the VAR(1)
component. The `predict()` method adds marginal prediction intervals for normal
margin fits:

```{r fitted-predict}
# VAR(1) fitted values: y_hat[t] = mu + Phi * (y[t-1] - mu)
fit_df <- fitted(fit)
head(fit_df)

# Predictions with 95% marginal intervals
pred_df <- predict(fit)
head(pred_df)

# Custom interval level
pred_90 <- predict(fit, ci_level = 0.90)
```

### Tidy Parameter Summary

Use `as.data.frame()` to export the full cmdstanr summary as a tidy data frame:

```{r as-data-frame}
param_df <- as.data.frame(fit)
head(param_df)
```

### Clinical Interpretation

`interpret_rho_trajectory()` provides a human-readable summary of the
estimated dependence trajectory:

```{r interpret}
interpret_rho_trajectory(fit)
```

## Visualisation

```{r plot}
# Rho trajectory with credible intervals
# Red line = true trajectory (simulation only)
plot_rho(fit, true_rho = sim$true_params$rho)

# VAR coefficient heatmap
plot_phi(fit)

# MCMC diagnostics (trace, Rhat, ESS)
plot(fit, type = "diagnostics")

# Posterior predictive check
# Currently available for normal and exponential margins
plot_ppc(fit)
```

## LOO-CV Model Comparison

Compare the DC-VAR against a constant-correlation baseline:

```{r compare}
fit_constant <- dcVar_constant(sim$Y_df, vars = c("y1", "y2"))

dcVar_compare(dcVar = fit, constant = fit_constant)
```
