---
title: "Comparing Copula VAR Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Copula VAR Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(Sys.getenv("NOT_CRAN"), "true")
has_cmdstan <- requireNamespace("cmdstanr", quietly = TRUE) &&
  !is.null(tryCatch(cmdstanr::cmdstan_path(), error = function(e) NULL))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = NOT_CRAN && has_cmdstan
)
```

## Overview

The dcVar package provides three copula VAR models:

1. **DC-VAR** (`dcVar()`): Time-varying correlation via a random walk on the Fisher-z scale. Best for smooth, gradual changes.

2. **HMM Copula** (`dcVar_hmm()`): Discrete regime-switching with K hidden states. Best for abrupt changes between distinct coupling regimes.

3. **Constant Copula** (`dcVar_constant()`): Single time-invariant correlation. Serves as a null/baseline model.

## Mathematical Background

For normal-margin fits, all three models share a common decomposition of the
bivariate likelihood into marginal and copula components. The package also
supports exponential, gamma, and skew-normal margins for these three models,
but the copula structure below is the common part.

### VAR(1) Marginals

Each variable follows a VAR(1) process with innovations:

$$y_t = \mu + \Phi \, (y_{t-1} - \mu) + \varepsilon_t, \qquad \varepsilon_t \sim N(0, \text{diag}(\sigma_{\varepsilon}^2))$$

The normal marginal densities provide the per-variable likelihoods in this
special case. For non-normal margins, dcVar uses the same VAR recursion and
Gaussian copula but swaps in the corresponding marginal likelihood and
transform.

### Gaussian Copula

The dependence between variables is captured by a Gaussian copula with
time-varying correlation $\rho_t$. The log-likelihood contribution of the
copula at time $t$ is:

$$\log c(u_{1t}, u_{2t}; \rho_t) = -\frac{1}{2}\log(1 - \rho_t^2) + \frac{\rho_t^2 \, (z_{1t}^2 + z_{2t}^2) - 2\rho_t \, z_{1t} z_{2t}}{2(1 - \rho_t^2)}$$

where $z_{it} = \Phi^{-1}(u_{it})$ and $u_{it} = \Phi(\varepsilon_{it} / \sigma_{\varepsilon,i})$.

### How the Models Differ

- **DC-VAR**: The correlation evolves as a random walk on the Fisher-z scale:
  $z_t = z_{t-1} + \omega_t$ with $\omega_t \sim N(0, \sigma_\omega^2)$, and
  $\rho_t = \tanh(z_t)$.

- **HMM Copula**: The correlation is state-dependent: $\rho_t = \rho_{s_t}$,
  where $s_t \in \{1, \ldots, K\}$ follows a Markov chain with transition
  matrix $A$. State-specific correlations use an ordered constraint
  ($\rho_1 < \rho_2 < \ldots$) to prevent label switching.

- **Constant Copula**: $\rho_t = \rho$ for all $t$. A special case of both
  the DC-VAR ($\sigma_\omega = 0$) and HMM ($K = 1$).

## Fitting All Three Models

```{r fit-all}
library(dcVar)

# Simulate data with a step-function trajectory
sim <- simulate_dcVar(
  T = 200,
  rho_trajectory = rho_step(200, rho_before = 0.7, rho_after = 0.3),
  seed = 42
)

# Fit all three models on the same data
fit_dc  <- dcVar(sim$Y_df, vars = c("y1", "y2"), seed = 1)
fit_hmm <- dcVar_hmm(sim$Y_df, vars = c("y1", "y2"), K = 2, seed = 2)
fit_con <- dcVar_constant(sim$Y_df, vars = c("y1", "y2"), seed = 3)
```

## LOO-CV Comparison

```{r compare}
dcVar_compare(dcVar = fit_dc, hmm = fit_hmm, constant = fit_con)
```

## Extracting Tidy Summaries

All fit objects support `as.data.frame()` for exporting tidy parameter
summaries. The three core time-series fits shown here also support
`fitted()`/`predict()` for one-step-ahead values; SEM and multilevel fits do
not currently implement those methods:

```{r tidy}
# Full parameter summary as a data frame
param_df <- as.data.frame(fit_dc)
head(param_df)

# Predictions with marginal intervals
pred_df <- predict(fit_hmm)
head(pred_df)
```

## HMM-Specific Outputs

The HMM model provides additional outputs related to state inference:

```{r hmm}
# State posteriors, Viterbi path, transition matrix
states <- hmm_states(fit_hmm)

# State-specific rho values
states$rho_state

# Transition matrix
states$A

# Visualise state posteriors
plot(fit_hmm, type = "states")

# Transition matrix heatmap
plot(fit_hmm, type = "transition")
```

## Clinical Interpretation

`interpret_rho_trajectory()` provides a model-aware narrative:

```{r interpret}
interpret_rho_trajectory(fit_dc)
interpret_rho_trajectory(fit_hmm)
interpret_rho_trajectory(fit_con)
```

## When to Use Which Model

- **Smooth dynamics** (therapy effects, developmental trends): Use DC-VAR
- **Abrupt regime switches** (treatment onset, relapse): Use HMM Copula
- **No time variation expected**: Use Constant Copula as baseline
- **Uncertain**: Fit all three and compare with LOO-CV
