---
title: "Simulation Study Tools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Study Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(Sys.getenv("NOT_CRAN"), "true")
has_cmdstan <- requireNamespace("cmdstanr", quietly = TRUE) &&
  !is.null(tryCatch(cmdstanr::cmdstan_path(), error = function(e) NULL))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = NOT_CRAN && has_cmdstan
)
```

## Trajectory Generators

dcVar provides several trajectory generators for simulation studies. All return
a numeric vector of length T-1 (rho values for time steps 2 through T):

```{r trajectories, eval = TRUE}
library(dcVar)

# Smooth trajectories
rho_constant(100, rho = 0.5)
rho_decreasing(100, rho_start = 0.7, rho_end = 0.3)
rho_increasing(100, rho_start = 0.3, rho_end = 0.7)
rho_random_walk(100, sigma_omega = 0.05, seed = 42)

# Step-function trajectories
rho_step(100, rho_before = 0.7, rho_after = 0.3, breakpoint = 0.5)
rho_double_step(100, rho_levels = c(0.7, 0.3, 0.7))

# Named scenarios
rho_scenario("double_relapse", T = 150)
```

### Visualising Trajectories

`plot_trajectories()` shows all built-in scenarios side by side:

```{r plot-trajectories, eval = TRUE}
# Compare all built-in scenarios
plot_trajectories(100)

# Subset of scenarios
plot_trajectories(100, scenarios = c("decreasing", "single_middle", "double_relapse"))
```

## Simulating Data

`simulate_dcVar()` generates bivariate VAR(1) data with a given rho trajectory
and known ground-truth parameters:

```{r simulate, eval = TRUE}
sim <- simulate_dcVar(
  T = 150,
  rho_trajectory = rho_decreasing(150),
  mu = c(0, 0),
  Phi = matrix(c(0.3, 0.1, 0.1, 0.3), 2, 2),
  sigma_eps = c(1, 1),
  seed = 42
)
```

### Breakpoint Data Simulator

`simulate_breakpoint_data()` provides a simplified interface for generating
data with single or double breakpoints:

```{r breakpoint, eval = TRUE}
# Single breakpoint
sim_bp <- simulate_breakpoint_data(T = 100, type = "single", seed = 42)
head(sim_bp$Y_df)
sim_bp$true_params$breakpoint

# Double breakpoint
sim_bp2 <- simulate_breakpoint_data(T = 100, type = "double", seed = 42)
```

## Data Preparation

Before fitting, data must be prepared with `prepare_dcVar_data()`,
`prepare_hmm_data()`, or `prepare_constant_data()`. These functions handle
standardisation, prior hyperparameters, and validation:

```{r prepare, eval = TRUE}
df <- data.frame(time = 1:100, y1 = rnorm(100), y2 = rnorm(100))

# DC-VAR data (includes sigma_omega prior)
stan_data <- prepare_dcVar_data(df, vars = c("y1", "y2"))
str(stan_data[c("T", "D", "sigma_mu_prior", "sigma_omega_prior")])

# HMM data (includes K, kappa, alpha_off)
hmm_data <- prepare_hmm_data(df, vars = c("y1", "y2"), K = 3)
str(hmm_data[c("K", "kappa", "alpha_off")])

# Constant data (simplest)
const_data <- prepare_constant_data(df, vars = c("y1", "y2"))
```

## Parameter Recovery Metrics

After fitting, evaluate recovery with:

```{r metrics-fit, include = FALSE}
fit <- dcVar(
  sim$Y_df,
  vars = c("y1", "y2"),
  chains = 2,
  iter_warmup = 250,
  iter_sampling = 250,
  seed = 99,
  refresh = 0
)
```

```{r metrics}
# Extract estimated rho
rho_df <- rho_trajectory(fit)

# Compute recovery metrics
metrics <- compute_rho_metrics(
  rho_true = sim$true_params$rho,
  rho_est = rho_df$mean,
  rho_lower = rho_df$q2.5,
  rho_upper = rho_df$q97.5
)

metrics$bias
metrics$coverage
metrics$correlation
```

## Running a Mini Simulation Study

```{r mini-study}
n_reps <- 5
results <- lapply(1:n_reps, function(i) {
  sim <- simulate_dcVar(
    T = 100,
    rho_trajectory = rho_decreasing(100),
    seed = i * 1000
  )

  fit <- dcVar(sim$Y_df, vars = c("y1", "y2"),
               chains = 2, iter_warmup = 500, iter_sampling = 1000,
               seed = i, refresh = 0)

  rho_df <- rho_trajectory(fit)

  list(rho = compute_rho_metrics(
    rho_true = sim$true_params$rho,
    rho_est = rho_df$mean,
    rho_lower = rho_df$q2.5,
    rho_upper = rho_df$q97.5
  ))
})

# Aggregate across replications
agg <- aggregate_metrics(results)
print(agg$rho)
```
